---
title: "index.qmd"
author: Maeve Grady and Leanne Chook
format: 
  html:
    self-contained: true
editor_options: 
  chunk_output_type: console
---

\vspace{0.1in}

## Visualization 1

```{r}
library(tidyverse)
library(readxl)
library(sf)
library(tigris)
library(rgeoboundaries)
library(readr)
library(ggrepel)
library(janitor)



```

\newpage

## Visualization 2

```{r}

```

\newpage

## Visualization 3: Bar Graph of Alcohol Consumption between Males and Females

```{r}
alcohol <- read_csv("alcohol.csv")

# select data that is needed and make adjustments 
alcohol2 <- alcohol %>%
  select(Location, Period, Dim1, FactValueNumeric, FactValueNumericLow, FactValueNumericHigh, ParentLocation) %>%
  rename("country" = Location,
         "year" = Period, 
         "average" = FactValueNumeric, 
         "min" = FactValueNumericLow,
         "max" = FactValueNumericHigh,
         "sex" = Dim1,
         "region" = ParentLocation) %>%
  filter(sex != "Both sexes")

# plot graph
alcohol2 %>%
  ggplot(aes(x = year,
             y = average)) +
  geom_col(aes(fill = sex)) +
  facet_wrap(~region) +

 # add the labels
  labs(x = "Year",
       y = "Total Alcohol Consumption per Capita\nin Litres of Pure Alcohol",
       caption = "Source: World Health Organisation", 
       title = "Total Alcohol Consumption Since 2000 in Different Regions", 
       subtitle = "Has there been a notable trend in alcohol consumption over the years?", 
       fill = "Gender"
  ) + 
  
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        axis.title = element_text(size = 8, face = "bold"),
        legend.title = element_text(size = 8, face = "bold"),
        legend.text = element_text(size = 8)) +
  
  scale_fill_manual(values = c("lightpink", "lightblue"))
    

```

This visualization tracks alcohol consumption trends in different regions around the world between males and females. The theory is that alcohol consumption has increased drastically in the past 10 years, however, the results proved otherwise. Alcohol consumption seems to be somewhat consistent since 2000, with males consuming much more alcohol and females in all regions.

\newpage

## Visualization 4: Pie chart of infant mortality causes 

```{r}
#import mortality data
mortality <- read_excel("mortality.xlsx") %>% clean_names()%>% mutate(value_string = as.numeric(value_string))

###selecting only 2019 data, and only data from each region
mort_19 <- mortality %>% filter(year == 2019, who_region != "Global")


###defining a blank theme
blank_theme <- theme_minimal()+
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  plot.title=element_text(size=14, face="bold")
  )


mort_19 %>% ggplot(mapping = aes(x = "" , y = value_string, fill = cause)) +
  geom_bar(width = 1, stat = "identity")+
  coord_polar("y", start= 0)+
  theme_minimal() +
  facet_wrap(~who_region)+ blank_theme +
  theme(axis.text.x=element_blank()) +
  labs(title = "Infant Mortality by cause", 
       subtitle = "Subdivided by World Health Organization region",
       caption = "Data source: World Health Organization, 2019")


```

This is a quick snapshot of what leading causes of infant mortality are accross the world. As is evident by these side-by side comparisons, the Americas, Europe and the Western Pacific experience relatively fewer infant mortalities overall compared to Afria and South East Asia. 