---
title: "index.qmd"
author: Maeve Grady and Leanne Chook
format: 
  html:
    self-contained: true
editor_options: 
  chunk_output_type: console
---

\vspace{0.1in}

## Assignment 04 

For this assignment, we decided to focus on global health, in particular, primary healthcare and non-communicable diseases. A leading cause of death around the world has been due to non-communicable diseases, which are also known as "silent" diseases that are caused by lifestyle choices, and are easily preventable with well-resourced and robust primary healthcare infrastructure.

\vspace{0.1in}

We wanted to use these data visualizations to explore if better policy is needed to increase government spending on primary healthcare and what are some areas that the spending should be focused on. 


## Visualization 1: Scatterplot of Primary Health Expenditures and GDP per Capita

```{r}
library(tidyverse)
library(readxl)
library(sf)
library(tigris)
library(rgeoboundaries)
library(readr)
library(ggrepel)
library(janitor)

options(scipen = 999)

# read in the datasets

health_expenditures <- read_excel("nha_indicators.xlsx")
gdppc <- read_excel("gdp.xls")

# make the expenditures dataset wider 
expenditures_wide <- health_expenditures %>%
  select(Countries, Indicators, '2018')

expenditures_wide <- pivot_wider(data = health_expenditures, 
                                 id_cols = Countries, 
                                 names_from = Indicators, 
                                 values_from = '2018'
)

# select only relevant columns 
gdppc2018 <- gdppc %>%
  select(Countries, '2018') %>%
  rename("gdp" = '2018')


# merge the datasets
expenditures <- merge(expenditures_wide, gdppc2018, 
                      by = 'Countries',
                      all.x = TRUE)

# rename variables for ease of coding
expenditures <- expenditures %>%
  rename("population" = `Population (in thousands)`,
         "primary_health" = `Primary Health Care (PHC) Expenditure per Capita in US$`)

# change variables from characters to numeric values 
expenditures <- expenditures %>% 
  mutate(population = as.numeric(population),
         primary_health = as.numeric(primary_health))

expenditures <- expenditures %>% 
  mutate(lngdp = log(gdp),
         population = population/1000)

# plot the graph
expenditures %>%
  ggplot(mapping =
           aes(x = primary_health, 
               y = lngdp, 
               size = population)) +
  geom_point(color = "green", 
             alpha = 0.5) +
  geom_text_repel(mapping = aes(label = Countries), size = 1.5) +

  # set the scale
  scale_size(range = c(1, 15)) +
  
  # add the labels
  labs(x = "Primary Healthcare Expenditure\nper Capita (US$)",
       y = "Natural log of GDP",
       caption = "Source: World Health Organisation and The World Bank", 
       title = "Primary Healthcare Expenditure based on GDP per capita", 
       subtitle = "Does GDP impact spending on primary healthcare?", 
       size = "Population\nin millions"
  ) + 
  
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        axis.title = element_text(size = 8, face = "bold"),
        legend.title = element_text(size = 8, face = "bold"),
        legend.text = element_text(size = 8))

```

The visualization shows that higher GDP (which has been transformed to natural log) does not necessarily amount to greater spending in primary healthcare. The Scandinavian countries who have higher GDP's tend to spend more on primary healthcare, but other high GDP countries like those in the Middle East spend very little. A majority of the countries are clustered towards the lower end of the expenditure spectrum.

It is also surprising to note that the United States did not have any data on primary healthcare expenditure, and is therefore left out of the graph. 

\newpage

## Visualization 2

```{r}

```

\newpage

## Visualization 3: Bar Graph of Alcohol Consumption between Males and Females

```{r}
alcohol <- read_csv("alcohol.csv")

# select data that is needed and make adjustments 
alcohol2 <- alcohol %>%
  select(Location, Period, Dim1, FactValueNumeric, FactValueNumericLow, FactValueNumericHigh, ParentLocation) %>%
  rename("country" = Location,
         "year" = Period, 
         "average" = FactValueNumeric, 
         "min" = FactValueNumericLow,
         "max" = FactValueNumericHigh,
         "sex" = Dim1,
         "region" = ParentLocation) %>%
  filter(sex != "Both sexes")

# plot graph
alcohol2 %>%
  ggplot(aes(x = year,
             y = average)) +
  geom_col(aes(fill = sex)) +
  facet_wrap(~region) +

 # add the labels
  labs(x = "Year",
       y = "Total Alcohol Consumption per Capita\nin Litres of Pure Alcohol",
       caption = "Source: World Health Organisation", 
       title = "Total Alcohol Consumption Since 2000 in Different Regions", 
       subtitle = "Has there been a notable trend in alcohol consumption over the years?", 
       fill = "Gender"
  ) + 
  
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        axis.title = element_text(size = 8, face = "bold"),
        legend.title = element_text(size = 8, face = "bold"),
        legend.text = element_text(size = 8)) +
  
  scale_fill_manual(values = c("lightpink", "lightblue"))
    

```

This visualization tracks alcohol consumption trends in different regions around the world between males and females. The theory is that alcohol consumption has increased drastically in the past 10 years, however, the results proved otherwise. Alcohol consumption seems to be somewhat consistent since 2000, with males consuming much more alcohol and females in all regions.

\newpage

## Visualization 4: Pie chart of infant mortality causes 

```{r}
#import mortality data
mortality <- read_excel("mortality.xlsx") %>% clean_names() %>% mutate(value_string = as.numeric(value_string))

###selecting only 2019 data, and only data from each region
mort_19 <- mortality %>% filter(year == 2019, who_region != "Global")


###defining a blank theme
blank_theme <- theme_minimal()+
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  plot.title=element_text(size=14, face="bold")
  )


mort_19 %>% ggplot(mapping = aes(x = "" , y = value_string, fill = cause)) +
  geom_bar(width = 1, stat = "identity")+
  coord_polar("y", start= 0)+
  theme_minimal() +
  facet_wrap(~who_region)+ blank_theme +
  theme(axis.text.x=element_blank()) +
  labs(title = "Infant Mortality by cause", 
       subtitle = "Subdivided by World Health Organization region",
       caption = "Data source: World Health Organization, 2019")


```

This is a quick snapshot of what leading causes of infant mortality are accross the world. As is evident by these side-by side comparisons, the Americas, Europe and the Western Pacific experience relatively fewer infant mortalities overall compared to Afria and South East Asia. 

